Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-01 forecastflows.tex
new additions from today huuuuge effort

% General scientific method we all learn
Quantitative science rely on a model-based framework to confront hypothesis and data, making some approximations \citep{}. The general scientific method stresses out that the research question should guide both the design of the experiment and the corresponding model building. The experiment should then be conducted---according to this specific design. Finally, the resulting experimental data should be used to inform our model and differentiate between hypotheses---hopefully answering our initial question. However, this is often an idealized view and does not reflect the complexity of ecology \citep{}.

% Divergences from this are common 
Divergences from this scientific method are common. One explanation is the lack of rigor, leading to questionable practices, such as retrospectively crafting hypothesis to explain the results of a model rather than testing a clear pre-defined question (data-driven analysis, Fig. 1). But the reality of ecological research is also a major driver of the divergences from the ideal scientific method. Many important questions often cannot be addressed by conducting experiments and replications, and we must often rely on existing datasets to have a large-scale and long-term perspective. % I feel something is missing
However, this does not explain the persistent flaws and lack of overall coherence in ecological modeling. Trend estimation still often relies on a one-way 'inference',  % need a better word? to stress out that there is no feedback, eah paper applied a slightly different model 
and most of the time is spent fitting the model to empirical data (Fig. 1). For forecasting, researchers focus on making predictions with complex models, but the steps of model building and parameterization are not very transparent and not clearly delineated (Fig. 1). The different parts of the model are often calibrated separately rather than as a whole, and some parameter values are just fixed based on experiments and expert knowledge.

% There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
% Moving along the data-model space
% \item stressing the need to think about model before study design
% \item advancing data simulation
[...]

% More details on the workflow:  walk through the different steps
[...]

@@

% Trends - outline current problem
%  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology. 
Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \cite{soule1999conserving,soule1991conservation}. The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}. Such trends drove a number of new subfields within ecology---some of which are now complete disciplines within themselves \citep[such as conservation biology,][]{soule1985conservation}---focused on these problems, and potential ecological solutions to them CITES. Yet, as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overharvesting, pollution only increase, and anthropogenic climate change now clearly driving species loss \citep{waller2017bramble}, so has the data and its complexity.

@@

% Trends - improvements!
Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates. Returning to the example of a global estimate of trends in vertebrate populations of species over time and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate? Or a need to also find which species are declining most, including those that may have poor or no data? From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data. Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model. For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled. From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematicall}.

% Forecasting - outline current problem
Forecasting is a broad field with a diverse range of methods. Here, we focus on process-based modeling, which is often considered the gold standard in ecology and beyond. Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration, where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature).

% Forecasting - potential improvments (do I say too much ideas)
Applying the workflow to process-based models is a key for opening the black box. It would serve as a guide through the successive steps of model development. In particular, incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing strong degeneracies in the model design. This approach would force researchers to begin with a simpler version of the model, providing a clear pathway to support---or reject---the additional complexity and new parameters along the iterative development of the model.
Data fitting would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights. The workflow could also refocus attention on the research question, defining a clear and limited context in which the model should apply.
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own. Finally, an universal workflow provides an opportunity to merge statistical and process-based approaches, integrating mechanistic knowledge and leveraging robust statistical approaches.

@@

The misuse of models and misinterpretation of their outcomes (including by the authors) have multiple origins. One cause, almost tautological, is the pressure to publish academic work, which can lower research standards. This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
This a growing concern, leading to increase in reproducibility and data sharing practices.
However, model development in ecology is still rarely transparent, which limits model understandability and prevents peers from properly identifying potential issues. A significant portion of scientific debate thus becomes lost in methodological considerations rather than advancing our ecological understanding.

>>>>

% General scientific method we all learn
Quantitative science rely on a model-based framework to confront hypothesis and data, making some approximations \citep{}. The general scientific method stresses out that the research question should guide both the design of the experiment and the corresponding model building. The experiment should then be conducted---according to this specific design. Finally, the resulting experimental data should be used to inform our model and differentiate between hypotheses---hopefully answering our initial question. However, this is often an idealized view.
% and does not reflect the complexity of ecology \citep{}. I already say it below

% Divergences from this are common 
Divergences from this scientific method are common. One explanation is the lack of rigor, leading to questionable practices, such as retrospectively crafting hypothesis to explain the results of a model rather than testing a clear pre-defined question (data-driven analysis, Fig. 1). But the reality of ecological research is also a major driver of the divergences from the ideal scientific method. Many important questions often cannot be addressed by conducting experiments and replications, and we must often rely on existing datasets to have a large-scale and long-term perspective \citep{Hilborn1997}. % I feel something is missing
However, this does not explain the persistent flaws and lack of overall coherence in ecological modeling. Trend estimation still often relies on a one-way 'inference',  % need a better word? to stress out that there is no feedback, eah paper applied a slightly different model 
and most of the time is spent fitting the model to empirical data (Fig. 1). For forecasting, researchers focus on making predictions with complex models, but the steps of model building and parameterization are not very transparent and not clearly delineated (Fig. 1). The different parts of the model are often calibrated separately rather than as a whole, and some parameter values are just fixed based on experiments and expert knowledge.

% There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
% Moving along the data-model space, step-by-step
% \item stressing the need to think about model before study design
% \item 
To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed to move along the data-model space, following a coherent sequence of steps. In particular, more efforts should be placed on model evaluation before incorporating any real data. This would force the modeler to acknowledge that some parameters might be non-identifiable and to reconsider the model structure. Similarly, it is essential to assess whether model predictions---once parameters are informed by data---are consistent with observations. The strength of such workflow lies in its flexibility, making it applicable to a wide range of modeling approaches, from simple trend analyses to more complex process-based models. At each step, the modeler need to critically examine its understanding of ecological processes, questioning previous assumptions, and explicitly acknowledge sources of uncertainty. This approach has the potential to enhance model interpretability and allow for a more transparent evaluation of model strengths and limitations. It also replace parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.

% More details on the workflow:  walk through the different steps
% the paragraph is a bit loooong
The first step of the workflow is to define an explicit research question and formulate some hypotheses. This involves making assumptions about the most influential drivers, within the specific context of our study. This step help us to think about the mechanisms that could generate the data we observe, including the observational error. Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and designed to answer our research question. The general idea is to start with a relatively simple model, that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. The next step is to simulate data that reflect our assumptions about the data structure and biases, by fixing parameter values to some values (which is straightforward if the parameters are interpretable). We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. Some issues might arise at this step. Structural degeneracies in the model might become apparent, revealing that some parameters are highly non-identifiable. This is the first feedback loop: if the model is not able to recover known parameter values, we must reconsider its structure. This way, we ensure the model works well on simulated data before incorporating real observations. We can then fit it to real data, to obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. The next step is a new data simulation step, this time using our fitted model to generate predictions. This retrodictive check allows model output to be compared to observations. Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic. This is a second feedback loop. We can refine the model to integrate the missing process and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake input data, such as phylogenetic structure or observational biases (e.g. unbalanced data). This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

@@

% Trends - outline current problem
%  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology. 
Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \citep{soule1999conserving,soule1991conservation}. The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}. Such trends drove a number of new subfields within ecology---some of which are now complete disciplines within themselves \citep[such as conservation biology,][]{soule1985conservation}---focused on these problems, and potential ecological solutions to them CITES. Yet, as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overharvesting, pollution only increase, and anthropogenic climate change now clearly driving species loss \citep{waller2017bramble}, so has the data and its complexity.

@@

% Trends - improvements!
Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates. Returning to the example of a global estimate of trends in vertebrate populations of species over time and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate? Or a need to also find which species are declining most, including those that may have poor or no data? From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data. Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model. For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled. From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.

% Forecasting - outline current problem
Ecological forecasting is a broad field with a diverse range of methods. Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond. Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not accidental, but rather a constraint imposed by their structure, where many parameters compensate for one another.

These limitations are central to current debates. In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty. For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong discrepancies exist between model projections \citep{}. A promising avenue is the simplification of models, avoiding over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where predictions lose their usefulness for decision makers \citep{Saltelli2020}.

% Forecasting - potential improvments (do I say too much ideas)
Applying the workflow to process-based models is a key for opening the black box. It would serve as a guide through the successive steps of model development. In particular, incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing strong degeneracies in the model design. This approach would force researchers to begin with a simpler version of the model, providing a clear pathway to support---or reject---the additional complexity and new parameters along the iterative development of the model. 
% => quantifying parameter uncertainty
% incorporate additional observations and their respective uncertainties
Model calibration would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights. 
% It would help properly take into account for any issues regarding non-identifiability.
The workflow could also refocus attention on the research question, defining a clear and limited context in which the model should apply.
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own. An universal workflow provides an opportunity to merge statistical and process-based approaches, integrating mechanistic knowledge and leveraging robust statistical approaches.

@@

The misuse of models and misinterpretation of their outcomes (including by the authors) have multiple origins. One cause, well known, is the pressure to publish academic work, which can lower research standards. This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
This a growing concern, leading to increase in reproducibility and data sharing practices.
However, model development in ecology is still rarely transparent, which limits model understandability and prevents peers from properly identifying potential issues. A significant portion of scientific debate thus becomes lost in methodological considerations rather than advancing our ecological understanding.

% better training! on BOTH estimation AND prediction
% learn how to simulate data will allow students to understad what is the link between the maths (eg a distribution) and the ecological processes that could be a l'oeuvre to generate the observations, make better links bewteen disciplines
% \item estimation: being aware of what is a parameter, and mention uncertainty propagation
% \item forecasting: should be a natural outcome, not a finality

% \item ML (short-time forecast), benchmarking models are probably useful but should not be the core of our scientific practice, not the spirit!\\
% $\rightarrow$ moving ecology in the right direction!
